NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE, PG and PS lipids

### 2017-03-09 Manuscript/manuscript.tex
Updates

\title{NMRlipids IV: Other than PC headgroups} %Title of paper

@@

\author{O. H. Samuli Ollila}
\email[]{samuli.ollila@helsinki.fi}
%\homepage[]{Your web page}
%\thanks{}
%\altaffiliation{}
\affiliation{University of Helsinki}

@@

Based on superficial reading, the conclusions in the literature are roughly \\ 
1) glycerol backbone structures are largely similar irrespectively of the headroup \cite{gally81}, \\
2) glycerol backbone and headgroup strucure and behaviour are similar in model membranes and in bacteria \cite{gally81,scherer87,seelig90}, \\
3) headgroup structures are similar in PC, PE and PG lipids, while headgroup is more rigid in PS lipids \cite{wohlgemuth80,buldt81}. \\
Extensive discussion about structural details of PE, PG or PS headgroups do not exists (as far as I know), 
In contrast to PC lipids (see \cite{botan15} and references therein).

@@

Comparison between experiments and simulations with CHARMM GUI
parameters for PS lipids in Fig. \ref{HGorderParametersPOPS} suggest that
the model overestimates magnitudes in $\beta$ and g$_3$ order parameters
and forking in $\alpha$.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPOPS.eps}
  \caption{\label{HGorderParametersPOPS}
    Order parameters for POPS headgroup and glycerol backbone from simulations and experiments without CaCl$_2$ 
    (DOPS from \cite{browning80} contains 0.1M of NaCl, POPC:POPS mixture from \cite{roux90}).
    Absolute values are shown for experimental data, because signs are not known.
    Simulations values are -S$_{CH}$
  }
  \todo{More simulation data for lipids with different headgroups to be collected}
  \todo{CHARMM GUI simulation contains only counter ions as potassium.
    All experiments here contain some amount of sodium salt. The best
    ion concentrations for comparison should be figured out.}
  \todo{Experimental signs of the order parameters would highly useful.}
\end{figure}

Comparison between experiments and simulations with CHARMM GUI
parameters for PG lipids in Fig. \ref{HGorderParametersPOPG} suggest that
the model overestimates forking in $\gamma$ and $\alpha$ order parameters,
and magnitudes of $\beta$ and $g_1$ order parameters.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPOPG.eps}
  \caption{\label{HGorderParametersPOPG}
    Order parameters for PG headgroup and glycerol backbone from simulations and experiments without CaCl$_2$ 
    (POPG from \cite{borle85} contains 10mM of PIPES, DPPG from \cite{wohlgemuth80} contains 10mM PIPES and 100mM CaCl,
    E.Coli PG results from \cite{gally81}).
    Absolute values are shown, because signs are not known for experimental data.
    Glycerol backbone order parameters are negative in simulations, while $alpha$, $\beta$ and
    smaller $\gamma$ are positive.
  }
  \todo{More simulation data for lipids with different headgroups to be collected}
  \todo{CHARMM GUI simulation contains only counter ions as potassium.
    All experiments here contain some amount of sodium salt. The best
    ion concentrations for comparison should be figured out.}
  \todo{Experimental signs of the order parameters would highly useful.}
\end{figure}

\section{Ca$2+$ binding in bilayers with negatively charged PG and PS lipids}

@@

PC headgroup order parameters increase when negatively charged
PS or PG are added to PC bilayer in the absense of added CaCl$_2$,
as expected based on electrometer concept \cite{seelig87}
(see Fig. \ref{OrderParametersWithCaClBELOW1M}).
%In electrometer concept this is explained by the tilting of
%headgroup more parallel to membrane normal \cite{??}.
Further, the order parameters decrease with the addition
of CaCl$_2$ and the decrease is more pronounced for systems with more
negatively charged lipids (see Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}).
At CaCl$_2$ concentrations ($\sim$ 50-300mM) where order parameters reach the values for pure PC,
the Ca2+ binding is presumably fully cancels the charge from negative lipids and
overcharging occurs above these concenterations.
The interpretation of this data and some other results has been that \cite{seelig90}
\begin{displayquote}
  {\it ''(i) Ca$^{2+}$ binds to neutral lipids (phosphatidylcholine, phosphatidylethanolamine) and negatively charged lipids
  (phosphatidylglycerol) with approximately the same binding constant of K = 10-20 M$^{-1}$; (ii) the free Ca$^{2+}$
  concentration at the membrane interface is distinctly enhanced if the membrane carries a negative surface
  charge, either due to protein or to lipid; (iii) increased inter-facial Ca$^{2+}$ also means increased amounts
  of bound Ca$^{2+}$ at neutral and charged lipids; (iv) the actual binding step can be described by a Langmuir
  adsorption isotherm with a 1 lipid:1 Ca$^{2+}$ stoichiometry, provided the interfacial concentration C$_M$, is
  used to describe the chemical binding equilibrium.''}
\end{displayquote}

>>>>

\title{NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE, PG and PS lipids} %Title of paper

@@

\author{O. H. Samuli Ollila}
\email[]{samuli.ollila@helsinki.fi}
%\homepage[]{Your web page}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\affiliation{Institute of Biotechnology, University of Helsinki}

@@

Based on superficial reading, the conclusions in the literature are roughly \\ 
1) glycerol backbone structures are largely similar irrespectively of the headroup \cite{gally81}, \\
2) glycerol backbone and headgroup structure and behaviour are similar in model membranes and in bacteria \cite{gally81,scherer87,seelig90}, \\
3) headgroup structures are similar in PC, PE and PG lipids, while headgroup is more rigid in PS lipids \cite{wohlgemuth80,buldt81}. \\
Extensive discussion about structural details of PE, PG or PS headgroups do not exists (as far as I know), 
In contrast to PC lipids (see \cite{botan15} and references therein).

@@

Comparison between experiments and simulations with CHARMM GUI
parameters for PS lipids in Fig. \ref{HGorderParametersPOPS} suggest that
the model overestimates magnitudes in $\beta$ and g$_3$ order parameters
and forking in $\alpha$.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPOPS.eps}
  \caption{\label{HGorderParametersPOPS}
    Order parameters for POPS headgroup and glycerol backbone from simulations with CHARMM GUI parameters and experiments without CaCl$_2$ 
    (DOPS from \cite{browning80} contains 0.1M of NaCl, POPC:POPS mixture from \cite{roux90}).
    Absolute values are shown for experimental data, because signs are not known.
    Simulations values are -S$_{CH}$
  }
  \todo{More simulation data for lipids with different headgroups to be collected}
  \todo{CHARMM GUI simulation contains only counter ions as potassium.
    All experiments here contain some amount of sodium salt. The best
    ion concentrations for comparison should be figured out.}
  \todo{Experimental signs of the order parameters would highly useful.}
\end{figure}

Comparison between experiments and simulations with CHARMM GUI
parameters for PG lipids in Fig. \ref{HGorderParametersPOPG} suggest that
the model overestimates forking in $\gamma$ and $\alpha$ order parameters,
and magnitudes of $\beta$ and $g_1$ order parameters.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPOPG.eps}
  \caption{\label{HGorderParametersPOPG}
    Order parameters for PG headgroup and glycerol backbone from simulations with CHARMM GUI parameters and experiments without CaCl$_2$ 
    (POPG from \cite{borle85} contains 10mM of PIPES, DPPG from \cite{wohlgemuth80} contains 10mM PIPES and 100mM CaCl,
    E.Coli PG results from \cite{gally81}).
    Absolute values are shown, because signs are not known for experimental data.
    Glycerol backbone order parameters are negative in simulations, while $alpha$, $\beta$ and
    smaller $\gamma$ are positive.
  }
  \todo{More simulation data for lipids with different headgroups to be collected}
  \todo{CHARMM GUI simulation contains only counter ions as potassium.
    All experiments here contain some amount of sodium salt. The best
    ion concentrations for comparison should be figured out.}
  \todo{Experimental signs of the order parameters would highly useful.}
\end{figure}

\section{Ca$^{2+}$ binding in bilayers with negatively charged PG and PS lipids}

@@

PC headgroup order parameters increase when negatively charged
PS or PG are added to PC bilayer in the absense of added CaCl$_2$,
as expected based on electrometer concept \cite{seelig87}
(see Fig. \ref{OrderParametersWithCaClBELOW1M}).
%In electrometer concept this is explained by the tilting of
%headgroup more parallel to membrane normal \cite{??}.
Further, the order parameters decrease with the addition
of CaCl$_2$ and the decrease is more pronounced for systems with more
negatively charged lipids (see Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}).
At CaCl$_2$ concentrations ($\sim$ 50-300mM) where order parameters reach the values for pure PC,
the Ca2+ binding presumably fully cancels the charge from negative lipids and
overcharging occurs above these concenterations.
The interpretation of this data and some other results has been that \cite{seelig90}
\begin{displayquote}
  {\it ''(i) Ca$^{2+}$ binds to neutral lipids (phosphatidylcholine, phosphatidylethanolamine) and negatively charged lipids
    (phosphatidylglycerol) with approximately the same binding constant of K = 10-20 M$^{-1}$; \\
    (ii) the free Ca$^{2+}$
    concentration at the membrane interface is distinctly enhanced if the membrane carries a negative surface
    charge, either due to protein or to lipid; \\
    (iii) increased inter-facial Ca$^{2+}$ also means increased amounts
    of bound Ca$^{2+}$ at neutral and charged lipids; \\
    (iv) the actual binding step can be described by a Langmuir
    adsorption isotherm with a 1 lipid:1 Ca$^{2+}$ stoichiometry, provided the interfacial concentration C$_M$, is
    used to describe the chemical binding equilibrium.''}
\end{displayquote}
